---
title: "A title for your project"
author: "Johann Wagner - u6958957"
date: "2023-10-02"
output:  
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
editor_options: 
  markdown: 
    wrap: 72
---

# Word/figure count

Words: [The number of words in your document, calculated using the word_count() function at the end of this document]
Figures: [The number of figures in your document. You can just count these]

# Location on GitHub

The root directory on GitHub: [DS4B-final-project](https://github.com/johann-wagner/DS4B-final-project/tree/main).

# Data Description

The data used in this project is from The Atlas of Living Australia (ALA), which is "a collaborative, digital and open infrastructure repository that aggregates biodiversity data from multiple sources" ([ALA, 2022](https://support.ala.org.au/support/solutions/articles/6000261410-introduction-to-the-ala)).

Specifically, this analysis uses species occurrence records for seven invasive animal species:

1.  [European Rabbits (Oryctolagus cuniculus)](https://bie.ala.org.au/species/https://biodiversity.org.au/afd/taxa/692effa3-b719-495f-a86f-ce89e2981652)

2.  [European Red Foxes (Vulpes vulpes)](https://bie.ala.org.au/species/https://biodiversity.org.au/afd/taxa/2869ce8a-8212-46c2-8327-dfb7fabb8296)

3.  [Cane Toads (Rhinella marina)](https://bie.ala.org.au/species/https://biodiversity.org.au/afd/taxa/e79179f8-44bf-46eb-b139-a2e66f6045c1)

4.  [Feral Cats (Felis catus)](https://bie.ala.org.au/species/https://biodiversity.org.au/afd/taxa/7de6b16a-f854-4b4b-88cf-81868ce74ad8)

5.  [Feral Horses (Equus (Equus) caballus)](https://bie.ala.org.au/species/https://biodiversity.org.au/afd/taxa/86b4395d-97a7-470d-b35b-306e01dfbab6)

6.  [Feral Pigs (Sus scrofa)](https://bie.ala.org.au/species/https://biodiversity.org.au/afd/taxa/948d8ca1-62fc-4d03-a79d-4be2de837264)

7.  [Red Imported Fire Ants (Solenopsis invicta)](https://bie.ala.org.au/species/https://biodiversity.org.au/afd/taxa/5b3fe1c6-1c2c-4e20-819f-2e6d8fa3d332)

I was interested in the spatial and temporal variables for all 7 species, particularly the longitude and latitude values (`decimalLongitude` and `decimalLatitude`) and date of recording (`eventDate`) for each record.

# Questions/Aims

## Initial Aims:

1.   Develop an RShiny app that showcases the spatial and temporal (monthly) occurrence of ~~the top 5-10~~ 7 invasive animal species in Australia by state/territory ~~and by regionality (urban vs. regional)~~

2.   ~~Visualise and calculate the proportion of the selected invasive animal species observed in an urban area~~

3.   Create a download option for users to download the cleaned dataset for the selected invasive animal species for a particular state/territory

## Justifications:

Initially, I was interested in the spatial distribution of magpies in Canberra as I am an avid cyclist and wanted to find the hotspots of magpie records and linking/statistically modelling this data with a collaborative magpie swooping database called [magpiealert.com](https://www.magpiealert.com/). Unfortunately, the data was not available and the sampling bias of the occurrence data may not be truely representative, as observations are more likely to be near roads, airports, and major urban areas ([Kellie, 2022](https://labs.ala.org.au/posts/2022-07-22_sample-bias/post.html?panelset=mantids)).

This made me interested in simply visualising the data spatially, instead of using any statistical methodologies and focusing the project on whether there would be major differences/patterns of occurrence records in urban vs. non-urban/regional areas. Unfortunately, given the scope of the project, calculating the urban vs. non-urban analysis would have been too much scope-creep, so this aim was ignored (maybe post-project exploration).

Hughes et al. ([2021](https://onlinelibrary.wiley.com/doi/10.1111/ecog.05926)) and Guerin et al. ([2016](https://doi.org/10.1371/journal.pone.0144779)) highlight the spatial sampling bias associated with biodiversity data suggesting that these biases must first be understood before applying any statistical modelling methods. Interestingly, Australian research by Piccolo et al. ([2020](https://www.nature.com/articles/s41598-020-66719-x)) show that reptile research locations are highly positively correlated with proximity to universities suggesting biodiversity research locations are closely associated by accessibility. Similarly, temporal biases can also influence results, as species distributions can change across time ([Boakes et al, 2010](https://doi.org/10.1371/journal.pbio.1000385)).

The Atlas of Living Australia already have a browser-based spatial visualisation tool, where a specific species can be selected and plotted on a map. However, there seemed to not be a functionality to compare both the spatial _and_ temporal (by month) variables.

This dashboard is the beginning data exploration step and in the essence of open and citizen-science, I wanted to make my cleaned data easily available through the dashboard. Data sharing allows for further analysis and reusability ([Tedersoo et al., 2021](https://www.nature.com/articles/s41597-021-00981-0), [Ramachandran et al., 2021](https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2020EA001562)). I wanted my dashboard to follow the FAIR Data Principles ([Wilkinson et al., 2016](https://www.nature.com/articles/sdata201618)).

I chose invasive animal species, because of the numerous rabbits I see whenever I go cycling in Canberra. The other 6 species were chosen arbitrary based off this [article](https://www.nespthreatenedspecies.edu.au/news-and-media/media-releases/australia-s-10-worst-invasive-species-study), which summarises the results from Kearney et al.'s paper ([2018](https://www.nespthreatenedspecies.edu.au/media/qile4abf/kearney-et-al-2018_the-threats-to-aus-species_postprint.pdf)).

A deeper understanding of my thought process is documented in GitHub Issue [`#4`](https://github.com/johann-wagner/DS4B-final-project/issues/4).

The analysis was based off this [ALALabs article](https://labs.ala.org.au/posts/2023-05-16_dingoes/post.html) ([Smith et al., 2023](https://labs.ala.org.au/posts/2023-05-16_dingoes/post.html)).

# Raw data

The raw data is available using the `galah` package. A user must use an ALA-registered email at [ala.org.au](ala.org.au) and use the `galah_config()` function. Once completed, the custom-function `load_galah_occurrence_data()` was used, where the input argument `species_name` was the aforementioned scientific names of the seven species.

```{r}
### Purpose:
# To load the species occurrence data of any given species within the Atlas
# of Living Australia database.
suppressPackageStartupMessages({
  library(galah)
})
galah_config(email = "johann.wagner@gmail.com")

load_galah_occurrence_data <- function(
  # character string of scientific name of species
  species_name
) {
  species_occurrence <- galah_call() |>
    
    # To conduct a search for species scientific name
    galah_identify(species_name) |> # Ref [1]
    
    # Pre-applied filter to ensure quality-assured data
    # the "ALA" profile is designed to exclude lower quality records Ref [2]
    galah_apply_profile(ALA) |> 
    atlas_occurrences(mint_doi = TRUE)
  
  return(species_occurrence)
}



# References:
# - [1] https://bie.ala.org.au/species/https://biodiversity.org.au/afd/taxa/692effa3-b719-495f-a86f-ce89e2981652
# - [2] https://galah.ala.org.au/R/reference/galah_apply_profile.html
```

The DOIs for each species:

1.  [European Rabbits (Oryctolagus cuniculus)](https://doi.org/10.26197/ala.d0da2524-838a-4540-a160-fba8088f4850)

2.  [European Red Foxes (Vulpes vulpes)](https://doi.org/10.26197/ala.5f700174-e92e-4376-a243-be05ffa93f46)

3.  [Cane Toads (Rhinella marina)](https://doi.org/10.26197/ala.d6d70c2c-00de-4f76-8ea6-598b9562eb64)

4.  [Feral Cats (Felis catus)](https://doi.org/10.26197/ala.57a58cdc-88c0-45c1-9260-bd17c6fc4b62)

5.  [Feral Horses (Equus (Equus) caballus)](https://doi.org/10.26197/ala.c3cebb21-c2e0-41d0-8b5d-f446c2914827)

6.  [Feral Pigs (Sus scrofa)](https://doi.org/10.26197/ala.952aa727-1915-4c3f-8399-ed940c24bfef)

7.  [Red Imported Fire Ants (Solenopsis invicta)](https://doi.org/10.26197/ala.74b5dd29-4287-473d-8391-9184c8c62554)

# Data wrangling

Because the data is already in tidy format, the seven datasets are binded into one `tidied_data.csv`, then filtered with `excluded_data.csv`. The excluded data includes any `NA` values in the `eventDate`, `decimalLongitude`, and `decimalLatitude` columns and also includes data not located within any of the [*State and Territories - 2021*](https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/access-and-downloads/digital-boundary-files) spatial boundaries provided by the Australian Bureau of Statistics (ABS). Further details of this process can be found in the [scripts folder](https://github.com/johann-wagner/DS4B-final-project/tree/main/scripts) and [data_cleaning.Rmd](https://github.com/johann-wagner/DS4B-final-project/blob/main/processed_data/data_cleaning.Rmd).

```{r}
suppressPackageStartupMessages({
  library(readr)
  library(tidyverse)
})
  
excluded_data <- read_csv("processed_data/excluded_data.csv")
tidied_data <- read_csv("processed_data/tidied_data.csv")

cleaned_data <- tidied_data %>% 
  anti_join(
    excluded_data,
    join_by(recordID)
  )
```

# Sanity checks

Let's use `skim()` to do some sanity checks.

```{r}
suppressPackageStartupMessages({
  library(skimr)
})
skim(cleaned_data)
```

There are 328,342 records with the same number of unique values in the `recordID` column indicating each record is distinct. There are 8 unique `state` values and 7 `simpleName` values, which makes sense. The latest `eventDate` value is 2023-10-09, which is not in the future.

Let's spatially showcase the data.

```{r}
suppressPackageStartupMessages({
  options(timeout = 1000)
  devtools::install_github("wfmackey/absmapsdata")
  # To easily access the Australian Bureau of Statistics (ABS) spatial structures
  ### https://github.com/wfmackey/absmapsdata
  library(absmapsdata)
  library(ggplot2)
})
state2021 %>% 
  # Only state/territories
  filter(!state_code_2021 %in% c("9", "Z")) %>% 
  ggplot() +
  
  # Ref [1]: Create background polygon of the ACT
  geom_sf(
    aes(geometry = geometry)
  ) +
  
  geom_point(
    data = cleaned_data,
    aes(
      x = decimalLongitude,
      y = decimalLatitude
    ),
    alpha = 0.1,
    size  = 0.7
  ) +
  
  coord_sf() +
  
  theme_minimal() +
  
  labs(title = "Sanity Checked! All data points are in Australia!")
```

Interestingly, it seems that there are a few observations that are to the east of the NSW border, which is likely an island part of NSW.

# Addressing the questions/aims

[This is the *main* section of your work. For example, if you are doing
an EDA, this is where you would summarise it. If you are building
something like a package or shiny app, a description of that package /
app and demonstration of its function(s) would go here. Use appropriate
sub-headings.]

# References

-   <https://support.ala.org.au/support/solutions/articles/6000261410-introduction-to-the-ala>
-   https://onlinelibrary.wiley.com/doi/10.1111/ecog.05926
-   https://labs.ala.org.au/posts/2022-07-22_sample-bias/post.html?panelset=mantids
https://doi.org/10.1371/journal.pone.0144779
- https://doi.org/10.1371/journal.pbio.1000385
https://www.nature.com/articles/s41598-020-66719-x
https://www.nature.com/articles/sdata201618

https://www.nature.com/articles/s41597-021-00981-0), https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2020EA001562

https://www.nespthreatenedspecies.edu.au/media/qile4abf/kearney-et-al-2018_the-threats-to-aus-species_postprint.pdf (DOI unavailable)
https://labs.ala.org.au/posts/2023-05-16_dingoes/post.html
- https://www.abs.gov.au/statistics/standards/australian-statistical-geography-standard-asgs-edition-3/jul2021-jun2026/access-and-downloads/digital-boundary-files